---
title: "01_data_pull"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(
  root.dir = "/Users/Jehan/Documents/FINLYTS-Group-2/Final_Paper_Latex/finlyts-ml-casestudy"
)
```

# PURPOSE: Pull all required data from Yahoo Finance and save to data/raw/
# ASSIGNED TO: Person 1 (Data Owner)

# REQUIRED SERIES:
# - Target stock:     [TICKER.PS]         (TBD - pending company selection)
# - Competitor stock: [TICKER.PS]         (TBD - pending company selection)
# - PSEi index:       ^PSEi
# - USD/PHP rate:     PHP=X
# - BTC:              BTC-USD
# - WTI Crude Oil:    CL=F
#
# DATE RANGE: 2019-01-01 to present

```{r main}
library(tidyquant)
library(tidyverse)
library(dplyr)
library(tidyr)
library(lubridate)
library(zoo)
```

```{r}
bpi <- read_csv("data/raw/bpi.csv")
bdo <- read_csv("data/raw/bdo.csv")
psei <- read_csv("data/raw/psei.csv")
```

# Cleaning the data

```{r}
bpi <- bpi %>%
  mutate(
    date = lubridate::mdy(Date)
  ) %>%
  mutate(
    bpi_price = as.numeric(bpi_price),
    bpi_volume = as.numeric(bpi_volume)
  ) %>%
  arrange(date)
```


```{r}
bdo <- bdo %>%
  mutate(
    date = lubridate::mdy(Date),
    Price = as.numeric(Price),
    Volume = as.numeric(gsub("[^0-9]", "", Vol.)),
    change_pct = as.numeric(gsub("%", "", `Change %`))
  ) %>%
  arrange(date)
```


```{r}
psei <- psei %>%
  mutate(
    date = lubridate::mdy(Date),
    Price = as.numeric(Price),
    Volume = ifelse(grepl("K", Vol.),
                    as.numeric(gsub("[^0-9.]", "", Vol.)) * 1000,
                    as.numeric(gsub("[^0-9.]", "", Vol.))),
    change_pct = as.numeric(gsub("%", "", `Change %`))
  ) %>%
  arrange(date)
```



# TODO: Align calendars and forward-fill PSE holidays
# TODO: Save raw series to data/raw/ as CSV

message("01_data_pull.R: Not yet implemented. Awaiting company selection.")
```
