---
title: "01_data_pull"
output: pdf_document
---

```{r setup, include=FALSE}
# Project configuration
# Ensure the 'here' package is used for portable file paths across different operating systems
if (!require("here")) install.packages("here")
here::i_am("R/01_data_pull.Rmd")

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 01 Data Pull & Cleaning
**Author:** Jehan & Aaron
**Purpose:** Load raw financial data from Investing.com, standardize formats, and merge into a master time series.

## 1. Libraries and Data Loading

We use `tidyverse` for general data manipulation and `here` for robust path management within the project structure.

```{r libraries}
library(tidyquant)
library(tidyverse)
library(lubridate)
library(here)
```

```{r load_data}
# Loading raw data files from the data/raw directory
# Investing.com CSV exports often contain commas in numeric fields and M/K suffixes for volume.
# These will be handled in the cleaning function below.
bpi_raw  <- read_csv(here("data", "raw", "bpi.csv"))
bdo_raw  <- read_csv(here("data", "raw", "bdo.csv"))
psei_raw <- read_csv(here("data", "raw", "psei.csv"))
exchange_rate_raw <- read_csv(here("data", "raw", "exchange rate.csv"))

# Pulling the U.S. 10-Year Treasury Yield (^TNX) from Yahoo Finance via tidyquant
# This serves as our second macro variable, the "global discount rate".
us10y_raw <- tq_get("^TNX", get = "stock.prices", from = "2019-01-01")
```

## 2. Cleaning Function

To maintain consistency between BPI, BDO, and PSEi datasets, we use a single unified cleaning function. This directly addresses the conversion issues where numeric columns with commas or unit suffixes (M/K) were previously being parsed as NA.

The function performs the following:
1. Standardizes column names to lowercase and removes non-alphanumeric characters.
2. Parses dates from MDY format.
3. Cleans numeric columns (Price, Open, High, Low) by removing commas before conversion.
4. Calculates Volume (Vol) by identifying 'M' (millions) or 'K' (thousands) suffixes.
5. Adds a prefix to the cleaned columns to prevent name collisions during the merge.

```{r cleaning_functions}
#' Clean financial data exported from Investing.com
#' @param df The raw data frame to be processed.
#' @param prefix A string prefix to append to each column except the date.
#' @return A processed data frame with standardized numeric columns.
clean_investing_data <- function(df, prefix = "") {
  # Standardize all column names to be snake_case and lowercase
  names(df) <- tolower(gsub("[^a-zA-Z0-9]", "_", names(df)))
  names(df) <- gsub("_+", "_", names(df))
  names(df) <- gsub("(^_|_$)", "", names(df))
  
  df_clean <- df %>%
    mutate(
      # Standardize the date column using lubridate mdy parsing
      date = mdy(date),
      
      # Convert price-related columns to numeric by stripping commas
      across(c(price, open, high, low), ~ as.numeric(gsub(",", "", .x))),
      
      # Transform volume strings into actual numerical values
      # We check for the suffix and multiply the parsed number accordingly.
      vol = case_when(
        grepl("M", vol) ~ as.numeric(gsub("[^0-9.]", "", vol)) * 1e6,
        grepl("K", vol) ~ as.numeric(gsub("[^0-9.]", "", vol)) * 1e3,
        TRUE ~ as.numeric(gsub("[^0-9.]", "", vol))
      ),
      
      # Remove percentage signs from the change column and convert to numeric
      change_pct = as.numeric(gsub("%", "", .data[["change"]]))
    ) %>%
    # Subset to the core required columns for analysis
    select(date, price, open, high, low, vol, change_pct) %>%
    arrange(date)
  
  # Append the prefix to columns if one is provided
  if (prefix != "") {
    df_clean <- df_clean %>%
      rename_with(~ paste0(prefix, "_", .x), -date)
  }
  
  return(df_clean)
}

#' Clean Yahoo Finance yield data
#' @param df The raw data frame from tq_get.
#' @param prefix A string prefix to append to the columns.
#' @return A processed data frame with standardized yield columns.
clean_yahoo_yield_data <- function(df, prefix = "us10y") {
  df %>%
    rename(price = close) %>% # ^TNX uses 'close' as the primary price point
    select(date, price, open, high, low, volume) %>%
    rename(vol = volume) %>%
    mutate(change_pct = (price / lag(price) - 1) * 100) %>%
    rename_with(~ paste0(prefix, "_", .x), -date) %>%
    arrange(date)
}
```

## 3. Execution and Merging

The individual stock and index series are cleaned and then merged into a single master data frame. 

We apply a forward-fill (Last Observation Carried Forward) strategy to handle dates where certain assets might not have traded (e.g., local holidays or weekends). This ensures a continuous series for time-series modeling.

```{r execute_cleaning}
# Processing each raw dataset individually
bpi_clean   <- clean_investing_data(bpi_raw,  prefix = "bpi")
bdo_clean   <- clean_investing_data(bdo_raw,  prefix = "bdo")
psei_clean  <- clean_investing_data(psei_raw, prefix = "psei")
fx_clean    <- clean_investing_data(exchange_rate_raw, prefix = "fx")
us10y_clean <- clean_yahoo_yield_data(us10y_raw, prefix = "us10y")

# Join all cleaned series together by the date column
master_data <- bpi_clean %>%
  full_join(bdo_clean,   by = "date") %>%
  full_join(psei_clean,  by = "date") %>%
  full_join(fx_clean,    by = "date") %>%
  full_join(us10y_clean, by = "date") %>%
  arrange(date)

# Filling missing data points from holidays or weekends
# This step is critical to maintain data integrity across different market schedules.
master_data <- master_data %>%
  fill(everything(), .direction = "down")

# Verification of cleaning results
# This prints the count of any remaining NA values to ensure the process was successful.
print("NA count per column after final cleaning and filling:")
colSums(is.na(master_data))
```

## 4. Complete Daily Time Index and Save

We expand the index to include every calendar day, then forward-fill gaps. 
This ensures a continuous series for time-series modeling and financial diagnostics.

```{r finalize_and_save}
# Create the complete calendar sequence from start to finish
full_dates <- tibble(
  date = seq(min(master_data$date), max(master_data$date), by = "day")
)

# Join the actual trading data back to the full calendar
# We fill down to carry the last available price through weekends/holidays.
master_data <- full_dates %>%
  left_join(master_data, by = "date") %>%
  arrange(date) %>%
  fill(everything(), .direction = "down")

# Creating the processed directory if it is not yet present
if (!dir.exists(here("data", "processed"))) dir.create(here("data", "processed"))

# Exporting the final master time-series data
write_csv(master_data, here("data", "processed", "master_series.csv"))
message("01_data_pull.R: Master series (full calendar) successfully saved to data/processed/master_series.csv")
```









