---
title: "02_feature_engineering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PURPOSE: Construct all mandatory features from rubric. NO LOOK-AHEAD BIAS.
# ASSIGNED TO: Person 2 (Feature Engineering and EDA)
# DEPENDS ON: 01_data_pull.R outputs in data/raw/

# FEATURE CHECKLIST (rubric Section 4):
# TREND: MA(5), MA(10), MA(20), EMA(12), EMA(26), MA ratio: MA5 / MA20
# MOMENTUM: Lagged returns (lag_1, lag_2, lag_3), Momentum 5-day, Momentum 10-day, RSI(14), MACD and signal line
# VOLATILITY: Rolling std dev (5-day, 10-day), Volatility ratio: vol5 / vol10
# MARKET AND CROSS-ASSET: PSEi returns, Competitor stock returns, USD/PHP returns
# TRANSFORMATIONS: Interaction term (momentum x volatility), Nonlinear (squared volatility)
# REGIME: COVID dummy
# TARGET VARIABLES: r_t1, r_t3, r_t5

# CRITICAL: All features must use only data available at time t or earlier.
# Use dplyr::lag() for all lookback windows.
# Use dplyr::lead() for targets only, assigned as separate columns.

```{r main}
library(tidyquant)
library(tidyverse)

message("02_feature_engineering.R: Not yet implemented.")
```
