---
title: "02_feature_engineering"
output: html_document
---

```{r setup, include=FALSE}
# Project configuration for feature engineering
if (!require("here")) install.packages("here")
here::i_am("R/02_feature_engineering.Rmd")

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 02 Feature Engineering & EDA
**Author:** Aaron (Feature Engineering and EDA)
**Purpose:** Construct all mandatory features from the rubric while strictly avoiding look-ahead bias.

## 1. Feature Checklist

As per rubric Section 4, the following features must be engineered from the master dataset:

### Trend Indicators
* Moving Averages: MA(5), MA(10), MA(20)
* Exponential Moving Averages: EMA(12), EMA(26)
* MA Ratio: MA(5) / MA(20)

### Momentum
* Lagged returns: lag_1, lag_2, lag_3
* Momentum: 5-day and 10-day windows
* Relative Strength Index: RSI(14)
* MACD: Moving Average Convergence Divergence and its associated signal line

### Volatility
* Rolling standard deviation: 5-day and 10-day windows
* Volatility ratio: vol(5) / vol(10)

### Market and Cross-Asset
* PSEi returns
* Competitor stock returns (e.g., BDO for BPI)
* Interaction terms (e.g., momentum multiplied by volatility)
* Nonlinear transformations (e.g., squared volatility)

### Target Variables
* Multi-horizon returns: r_t+1, r_t+3, r_t+5

## 2. Methodology and Implementation

The implementation must strictly adhere to time-series discipline:
1. All features must use only data available at time *t* or earlier to prevent look-ahead bias.
2. Use `dplyr::lag()` for all lookback windows and technical indicators.
3. Use `dplyr::lead()` only for target variables (labels), which are assigned to separate columns for supervised learning.

```{r libraries}
library(tidyquant)
library(tidyverse)
library(here)
```

```{r execute_feature_engineering}
# This section will load data/processed/master_series.csv and apply transformations.
# Currently awaiting implementation of technical indicators.
message("02_feature_engineering.Rmd: Feature construction script is ready for implementation.")
```
