---
title: "03_ols_baseline"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PURPOSE: Estimate OLS benchmark model for all three horizons (t+1, t+3, t+5)
# ASSIGNED TO: Person 3 (Modeling and Results)
# DEPENDS ON: 02_feature_engineering.R output in data/processed/

# REQUIREMENTS:
# - Time-series split: first 70-80% training, remainder test
# - Random splitting is NOT allowed
# - Naive benchmark: previous return (lag_1) as prediction
# - Output: RMSE, MAE, R-squared, Directional Accuracy per horizon
# - Save coefficient table as LaTeX to paper/tables/ols_coefficients.tex

# ADVICE FOR CARLOS:
# 1. Split Consistency: Use the same 80/20 chronological split defined in Section 10 of 02_feature_engineering.Rmd (cutoff ~April 2024) to ensure model comparability.
# 2. Multicollinearity: The feature set includes both 'ma_ratio' and raw MAs (ma5, ma10, ma20). OLS will suffer from high VIF. Conduct a VIF check and consider dropping the raw MAs in favor of the ratio for the linear baseline.

```{r main}
library(tidymodels)
library(tidyverse)

message("03_ols_baseline.R: Not yet implemented.")
```
