---
title: "03_ols_baseline"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PURPOSE: Estimate OLS benchmark model for all three horizons (t+1, t+3, t+5)
# ASSIGNED TO: Person 3 (Modeling and Results)
# DEPENDS ON: 02_feature_engineering.R output in data/processed/

# REQUIREMENTS:
# - Time-series split: first 70-80% training, remainder test
# - Random splitting is NOT allowed
# - Naive benchmark: previous return (lag_1) as prediction
# - Output: RMSE, MAE, R-squared, Directional Accuracy per horizon
# - Save coefficient table as LaTeX to paper/tables/ols_coefficients.tex

# ADVICE FOR CARLOS:
# 1. Split Consistency: Use the same 80/20 chronological split defined in Section 10 of 02_feature_engineering.Rmd (cutoff ~April 2024) to ensure model comparability.
# 2. Multicollinearity: The feature set includes both 'ma_ratio' and raw MAs (ma5, ma10, ma20). OLS will suffer from high VIF. Conduct a VIF check and consider dropping the raw MAs in favor of the ratio for the linear baseline.

```{r main}
library(tidymodels)
library(tidyverse)
library(tidyquant)
library(broom)

```{r data}
master_file <- read.csv("data/processed/master_series.csv", stringsAsfactors = "FALSE") 
```

```{r log returns}
df <- master |>
  mutate(
    r_bpi   <- log(bpi_price   / lag(bpi_price)),   # target asset
    r_bdo   <- log(bdo_price   / lag(bdo_price)),   # sector competitor
    r_psei  <- log(psei_price  / lag(psei_price)),  # market index
    r_fx    <- log(fx_price    / lag(fx_price)),    # USD/PHP macro
    r_us10y <- log(us10y_price / lag(us10y_price))  # US 10Y macro
  )
```

```{r split}
set.seed(42)
# Strict time-series split: first 80% = training, last 20% = test
n_total <- nrow(features)
n_train  <- floor(0.80 * n_total)
n_test   <- n_total - n_train

```



message("03_ols_baseline.R: Not yet implemented.")
```
