---
title: "03_ols_baseline"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PURPOSE: Estimate OLS benchmark model for all three horizons (t+1, t+3, t+5)
# ASSIGNED TO: Person 3 (Modeling and Results)
# DEPENDS ON: 02_feature_engineering.R output in data/processed/

# REQUIREMENTS:
# - Time-series split: first 70-80% training, remainder test
# - Random splitting is NOT allowed
# - Naive benchmark: previous return (lag_1) as prediction
# - Output: RMSE, MAE, R-squared, Directional Accuracy per horizon
# - Save coefficient table as LaTeX to paper/tables/ols_coefficients.tex

# ADVICE FOR CARLOS:
# 1. Split Consistency: Use the same 80/20 chronological split defined in Section 10 of 02_feature_engineering.Rmd (cutoff ~April 2024) to ensure model comparability.
# 2. Multicollinearity: The feature set includes both 'ma_ratio' and raw MAs (ma5, ma10, ma20). OLS will suffer from high VIF. Conduct a VIF check and consider dropping the raw MAs in favor of the ratio for the linear baseline.

```{r main}
library(tidyquant)     # SMA, EMA, RSI, MACD, runSD via TTR
library(tidyverse)
library(car)           # vif()
library(broom)         # tidy(), glance()
library(knitr)         # kable()
library(kableExtra)
```

```{r data}
master_file <- read.csv("data/processed/master_series.csv", stringsAsfactors = "FALSE")
cat("Rows:", nrow(master), "| Columns:", ncol(master), "\n")
cat("Date range:", format(min(master$date), "%Y-%m-%d"),
    "to", format(max(master$date), "%Y-%m-%d"), "\n")
```

```{r log_returns}
df <- master |>
  mutate(
    r_bpi   <- log(bpi_price   / lag(bpi_price)),   # target asset
    r_bdo   <- log(bdo_price   / lag(bdo_price)),   # sector competitor
    r_psei  <- log(psei_price  / lag(psei_price)),  # market index
    r_fx    <- log(fx_price    / lag(fx_price)),    # USD/PHP macro
    r_us10y <- log(us10y_price / lag(us10y_price))  # US 10Y macro
  )
```

```{r targets}
df <- df |> mutate (target_1d = lead(r_bpi, 1), 
        target_3d = lead(r_bpi, 1) + lead(r_bpi, 2) + lead(r_bpi, 3), 
        target_5d = lead(r_bpi, 1) + lead(r_bpi, 2) + lead(r_bpi, 3) +
                lead(r_bpi, 4) + lead(r_bpi, 5) 
```

```{r trend}
df <- df |>
  mutate(
    ma5      = SMA(bpi_price, n = 5),
    ma10     = SMA(bpi_price, n = 10),
    ma20     = SMA(bpi_price, n = 20),
    ema12    = EMA(bpi_price, n = 12),
    ema26    = EMA(bpi_price, n = 26),
    ma_ratio = ma5 / ma20          # short-run vs long-run trend divergence
  )
```

```{r momentum}
df <- df |>
  mutate(
    lag1  = lag(r_bpi, 1),
    lag2  = lag(r_bpi, 2),
    lag3  = lag(r_bpi, 3),
    mom5  = r_bpi - lag(r_bpi, 5),
    mom10 = r_bpi - lag(r_bpi, 10)
  ) |>
  mutate(rsi14 = RSI(bpi_price, n = 14))

macd_matrix <- MACD(df$bpi_price, nFast = 12, nSlow = 26, nSig = 9, maType = "EMA")
df <- df |>
  mutate(
    macd        = as.numeric(macd_matrix[, "macd"]),
    macd_signal = as.numeric(macd_matrix[, "signal"])
  )
```



message("03_ols_baseline.R: Not yet implemented.")
```
