---
title: "05_models_svm_ann"
output: html_document
---

```{r setup, include=FALSE}
# Install any missing packages before loading
pkgs <- c("tidymodels", "kernlab", "nnet", "kableExtra", "tidyverse", "here")
invisible(lapply(pkgs, function(p) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, quiet = TRUE)
}))
here::i_am("R/05_models_svm_ann.Rmd")

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# 05 Support Vector Machine and Artificial Neural Network

**Author:** Aaron
**Purpose:** Estimate Support Vector Machine (SVM, RBF kernel) and Artificial Neural Network (ANN) models across all three prediction horizons (1-day, 3-day, 5-day). Both models use feature normalisation via `step_normalize()` inside the tidymodels workflow, ensuring the scaling transform is fitted on the training fold only and never contaminates the assessment fold or test set.

**Depends on:** `data/processed/features.csv` (output of `02_feature_engineering.Rmd`)

**Note on ANN architecture:** The `nnet` package (as specified in the rubric) implements a single hidden-layer perceptron with logistic activation and L2 weight decay (`penalty`). It does **not** support ReLU activation or dropout. This is a constraint of the `nnet` package. The regularisation via weight decay serves a function analogous to dropout by penalising large weights and reducing overfitting.

---

## 1. Libraries

```{r libraries}
library(tidymodels)    # parsnip, recipes, tune, rsample
library(kernlab)       # SVM engine
library(nnet)          # single-layer ANN engine
library(kableExtra)    # LaTeX table export
library(tidyverse)
library(here)
```

---

## 2. Constants

```{r constants}
# Reproducibility
RANDOM_SEED <- 42L

# Data split — must mirror 02_feature_engineering.Rmd Section 10
TRAIN_PROP  <- 0.80

# Rolling-origin CV
CV_INITIAL  <- 200L   # minimum training window
CV_ASSESS   <- 60L    # assessment window (~one trading quarter)
CV_SKIP     <- 100L   # gives ~13 slices on 80% of 1850 obs; kernlab SVM is slow, increase to speed up

# SVM
SVM_GRID_SIZE <- 15L  # Latin hypercube grid points for cost x rbf_sigma

# ANN
ANN_EPOCHS     <- 200L
ANN_GRID_SIZE  <- 15L  # Latin hypercube grid points for hidden_units x penalty
ANN_HIDDEN_MIN <- 3L
ANN_HIDDEN_MAX <- 15L  # cap avoids overly large networks for n~1400 training obs

# Column definitions
TARGET_COLS  <- c("target_1d", "target_3d", "target_5d")
HORIZON_LABS <- c("1d",        "3d",        "5d")

PREDICTOR_COLS <- c(
  # contemporaneous returns (known at time t)
  "r_bpi", "r_bdo", "r_psei", "r_fx", "r_us10y",
  # trend
  "ma5", "ma10", "ma20", "ema12", "ema26", "ma_ratio",
  # momentum
  "lag1", "lag2", "lag3", "mom5", "mom10", "rsi14", "macd", "macd_signal",
  # volatility
  "vol5", "vol10", "vol20", "vol_ratio",
  # cross-asset
  "lag_psei",
  # engineered transformations
  "interact", "vol5_sq",
  # regime
  "hi_vol"
)
```

---

## 3. Load Data and Reproduce Split

```{r load_split}
features <- read_csv(
  here("data", "processed", "features.csv"),
  show_col_types = FALSE
)

n_total <- nrow(features)
n_train <- floor(TRAIN_PROP * n_total)
n_test  <- n_total - n_train

train_df <- features[seq_len(n_train), ]
test_df  <- features[(n_train + 1L):n_total, ]

cat("Total clean observations:", n_total, "\n")
cat("Training set            :", n_train,
    "| up to", format(train_df$date[nrow(train_df)], "%Y-%m-%d"), "\n")
cat("Test set                :", n_test,
    "| from", format(test_df$date[1], "%Y-%m-%d"), "\n")
```

---

## 4. Metrics Helper

```{r metrics_fn}
compute_metrics <- function(actual, predicted, naive_pred, model, horizon) {
  ss_res   <- sum((actual - predicted)^2)
  ss_total <- sum((actual - mean(actual))^2)

  tibble(
    Model    = model,
    Horizon  = horizon,
    RMSE     = sqrt(mean((actual - predicted)^2)),
    MAE      = mean(abs(actual - predicted)),
    R2       = 1 - ss_res / ss_total,   # can be negative out-of-sample
    DA       = mean(sign(predicted) == sign(actual)),
    Naive_DA = mean(sign(naive_pred) == sign(actual))
  )
}
```

---

## 5. Support Vector Machine (RBF Kernel)

SVM requires all numeric features to be on a comparable scale. The recipe includes `step_normalize(all_numeric_predictors())`, which centres and scales each predictor to zero mean and unit variance. Critically, `prep()` is called once per CV fold inside the tidymodels workflow — the normalisation parameters are derived only from the training portion of each fold and are never exposed to the assessment or test data.

We tune two kernel hyperparameters:

| Parameter | Controls |
|---|---|
| `cost` | Regularisation; lower values = wider margin, more slack |
| `rbf_sigma` | RBF kernel bandwidth; higher values = tighter decision boundary |

```{r svm_models}
set.seed(RANDOM_SEED)

svm_metrics_list <- vector("list", length(TARGET_COLS))

for (h in seq_along(TARGET_COLS)) {

  target_col    <- TARGET_COLS[h]
  horizon_label <- HORIZON_LABS[h]

  cat("\nSVM |", toupper(horizon_label), "horizon\n")

  train_h <- train_df %>% select(all_of(c(target_col, PREDICTOR_COLS)))
  test_h  <- test_df  %>% select(all_of(c(target_col, PREDICTOR_COLS)))

  cv_h <- rolling_origin(
    data       = train_h,
    initial    = CV_INITIAL,
    assess     = CV_ASSESS,
    skip       = CV_SKIP,
    cumulative = TRUE
  )
  cat("CV slices:", nrow(cv_h), "\n")

  # step_normalize fits mean and SD on the training partition of each CV fold only.
  # Fitting the scaler on the full training set before CV would leak fold-assessment info.
  svm_recipe <- recipe(
    as.formula(paste(target_col, "~ .")),
    data = train_h
  ) %>%
    step_normalize(all_numeric_predictors())

  svm_spec <- svm_rbf(
    cost      = tune(),
    rbf_sigma = tune()
  ) %>%
    set_engine("kernlab") %>%
    set_mode("regression")

  svm_wf <- workflow() %>%
    add_recipe(svm_recipe) %>%
    add_model(svm_spec)

  svm_grid <- grid_latin_hypercube(
    cost(),
    rbf_sigma(),
    size = SVM_GRID_SIZE
  )

  svm_tune <- tune_grid(
    svm_wf,
    resamples = cv_h,
    grid      = svm_grid,
    metrics   = metric_set(rmse, mae, rsq),
    control   = control_grid(verbose = FALSE, save_pred = FALSE)
  )

  best_svm <- select_best(svm_tune, metric = "rmse")
  cat("Best cost:", round(best_svm$cost, 5),
      "| rbf_sigma:", round(best_svm$rbf_sigma, 6), "\n")

  final_svm_wf  <- finalize_workflow(svm_wf, best_svm)
  final_svm_fit <- fit(final_svm_wf, data = train_h)

  svm_preds   <- predict(final_svm_fit, new_data = test_h)$.pred
  naive_preds <- test_df$lag1

  svm_metrics_list[[h]] <- compute_metrics(
    actual     = test_h[[target_col]],
    predicted  = svm_preds,
    naive_pred = naive_preds,
    model      = "SVM",
    horizon    = horizon_label
  )
  cat("RMSE:", round(svm_metrics_list[[h]]$RMSE, 6),
      "| DA:",  round(svm_metrics_list[[h]]$DA,   4),
      "| Naive DA:", round(svm_metrics_list[[h]]$Naive_DA, 4), "\n")

  saveRDS(
    final_svm_fit,
    here("output", "models", paste0("svm_", horizon_label, ".rds"))
  )
  cat("Model saved: output/models/svm_", horizon_label, ".rds\n", sep = "")
}

svm_metrics <- bind_rows(svm_metrics_list)
```

---

## 6. Artificial Neural Network (nnet — Single Hidden Layer)

The ANN uses `nnet` via tidymodels' `mlp()` interface. Architecture: one hidden layer with a tunable number of hidden units; linear activation on the output node (regression mode). Feature normalisation is applied via the same `step_normalize()` recipe as SVM, ensuring stable gradient descent and comparable input magnitudes.

We tune two parameters:

| Parameter | Controls |
|---|---|
| `hidden_units` | Width of the single hidden layer |
| `penalty` | L2 weight decay; prevents overfitting |

**Architectural note:** `nnet` uses logistic (sigmoid) activation on hidden units and does not natively support ReLU activation or dropout. The `penalty` (weight decay) parameter provides regularisation analogous to dropout by penalising large weight magnitudes.

```{r ann_models}
set.seed(RANDOM_SEED)

ann_metrics_list <- vector("list", length(TARGET_COLS))

for (h in seq_along(TARGET_COLS)) {

  target_col    <- TARGET_COLS[h]
  horizon_label <- HORIZON_LABS[h]

  cat("\nANN |", toupper(horizon_label), "horizon\n")

  train_h <- train_df %>% select(all_of(c(target_col, PREDICTOR_COLS)))
  test_h  <- test_df  %>% select(all_of(c(target_col, PREDICTOR_COLS)))

  cv_h <- rolling_origin(
    data       = train_h,
    initial    = CV_INITIAL,
    assess     = CV_ASSESS,
    skip       = CV_SKIP,
    cumulative = TRUE
  )
  cat("CV slices:", nrow(cv_h), "\n")

  # Normalisation ensures stable gradient descent across features with different scales
  ann_recipe <- recipe(
    as.formula(paste(target_col, "~ .")),
    data = train_h
  ) %>%
    step_normalize(all_numeric_predictors())

  ann_spec <- mlp(
    hidden_units = tune(),
    penalty      = tune(),
    epochs       = ANN_EPOCHS
  ) %>%
    set_engine("nnet") %>%
    set_mode("regression")

  ann_wf <- workflow() %>%
    add_recipe(ann_recipe) %>%
    add_model(ann_spec)

  ann_grid <- grid_latin_hypercube(
    hidden_units(range = c(ANN_HIDDEN_MIN, ANN_HIDDEN_MAX)),
    penalty(range = c(-5, 0)),   # log10 scale: 1e-5 to 1.0
    size = ANN_GRID_SIZE
  )

  ann_tune <- tune_grid(
    ann_wf,
    resamples = cv_h,
    grid      = ann_grid,
    metrics   = metric_set(rmse, mae, rsq),
    control   = control_grid(verbose = FALSE, save_pred = FALSE)
  )

  best_ann <- select_best(ann_tune, metric = "rmse")
  cat("Best hidden_units:", best_ann$hidden_units,
      "| penalty:", round(best_ann$penalty, 6), "\n")

  final_ann_wf  <- finalize_workflow(ann_wf, best_ann)
  final_ann_fit <- fit(final_ann_wf, data = train_h)

  ann_preds   <- predict(final_ann_fit, new_data = test_h)$.pred
  naive_preds <- test_df$lag1

  ann_metrics_list[[h]] <- compute_metrics(
    actual     = test_h[[target_col]],
    predicted  = ann_preds,
    naive_pred = naive_preds,
    model      = "ANN",
    horizon    = horizon_label
  )
  cat("RMSE:", round(ann_metrics_list[[h]]$RMSE, 6),
      "| DA:",  round(ann_metrics_list[[h]]$DA,   4),
      "| Naive DA:", round(ann_metrics_list[[h]]$Naive_DA, 4), "\n")

  saveRDS(
    final_ann_fit,
    here("output", "models", paste0("ann_", horizon_label, ".rds"))
  )
  cat("Model saved: output/models/ann_", horizon_label, ".rds\n", sep = "")
}

ann_metrics <- bind_rows(ann_metrics_list)
```

---

## 7. Results Summary

```{r results_table}
all_metrics <- bind_rows(svm_metrics, ann_metrics) %>%
  mutate(across(where(is.double), ~ round(., 5)))

kable(
  all_metrics,
  caption   = "SVM and ANN: Out-of-Sample Performance by Horizon",
  col.names = c("Model", "Horizon", "RMSE", "MAE", "R2", "DA", "Naive DA"),
  digits    = 5
) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

---

## 8. Save Outputs

```{r save_outputs}
write_csv(all_metrics, here("output", "predictions", "svm_ann_metrics.csv"))
message("Metrics saved to output/predictions/svm_ann_metrics.csv")

if (!dir.exists(here("paper", "tables"))) dir.create(here("paper", "tables"))

latex_tbl <- kable(
  all_metrics,
  format    = "latex",
  booktabs  = TRUE,
  escape    = TRUE,
  linesep   = "",
  caption   = "SVM and ANN: Out-of-Sample Performance by Horizon",
  label     = "svm_ann_metrics",
  col.names = c("Model", "Horizon", "RMSE", "MAE", "R2", "DA", "Naive DA"),
  digits    = 5
) %>%
  kable_styling(latex_options = c("hold_position", "scale_down")) %>%
  footnote(
    general           = paste0(
      "DA = Directional Accuracy (fraction of test-set observations where the ",
      "predicted sign matches the realised sign). ",
      "Naive DA uses the previous day's return (lag1) as the prediction. ",
      "R2 is out-of-sample and may be negative if the model performs worse than the mean. ",
      "SVM features are normalised inside the workflow (zero mean, unit variance). ",
      "ANN uses logistic hidden-unit activation and L2 weight decay via the nnet engine."
    ),
    general_title     = "Note.",
    footnote_as_chunk = TRUE,
    escape            = TRUE
  )

writeLines(
  as.character(latex_tbl),
  here("paper", "tables", "svm_ann_metrics.tex")
)
message("LaTeX table saved to paper/tables/svm_ann_metrics.tex")
message("05_models_svm_ann.Rmd: Complete.")
```
